How wise is the crowd: Can we infer people are accurate and competent merely because they agree with each other?

### 2024-11-24 paper.tex
update some p-value reporting after proofs, removing "=" when smaller than 0.001

\caption{Results of the analytic argument. The figure shows the average estimated competence and accuracy for a focal individual, depending on the proportion of individuals agreeing with them, in a categorical choice scenario. Here, we assume that 20 individuals answer a 5-choice question. For values higher than four, we assume that the focal individual is part of the majority. Below four, we assume that the focal individual is part of a minority, and that the majority size is four. Two situations are represented: one in which there is low variance in the distribution of competence in the population (orange line in the insert), and one in which there is high variance in this distribution (blue line in the insert).}\label{fig:figure-Benoit-2}
\end{figure}

@@

To account for dependencies of observations due to our within-participant design, we ran mixed models, with a random intercept and a random slopes of convergence for participants. In the models for our hypotheses, we control for the number of estimates provided to the participants (three or ten). Visualizations and descriptive statistics can be found in ESM, section \ref{exp1}. We find a positive effect of convergence on accuracy: Participants were more confident about their estimate in convergent scenarios (mean = 4.56, sd = 1.45) than in divergent ones (mean = 3.19, sd = 1.39; \(\hat{b}_{\text{Accuracy}}\) = 1.37 {[}1.225, 1.51{]}, p = \textless{} .001). We also find a positive effect of convergence on competence: participants rated players as more competent in convergent scenarios (mean = 4.75, sd = 1.24) than in divergent ones (mean = 3.52, sd = 1.27; \(\hat{b}_{\text{Competence}}\) = 1.23 {[}1.065, 1.4{]}, p = \textless{} .001).

In an exploratory, non-preregistered analysis, we tested whether the effect of convergence is larger on accuracy than on competence. To this end, we regressed the outcome score on convergence and its interaction with a binary variable indicating which outcome was asked for (accuracy or competence), while controlling for the number of informants. We do not find a statistically significant interaction that would indicate a difference of the effect of convergence (\(\hat{b}\) = 0.14 {[}-0.001, 0.271{]}, 0.052). Pooled across divergent and convergent conditions, we find that participants reported lower perceived accuracy than competence (\(\hat{b}\) = -0.26 {[}-0.359, -0.156{]}, \textless{} .001)

@@

To account for dependencies of observations due to our within-participant design, we ran mixed models, with a random intercept and a random slope for convergence for participants. We find evidence for all four hypotheses. As for the first set of hypotheses, to match the setting of experiment one, we reduced the sample of Experiment 3 to half of the participants, namely those who were assigned to the independence condition. On this reduced sample, we ran the exact same analyses as in Experiment 1 and replicated the results. As for accuracy, participants rated informants in convergent scenarios (mean = 5.28, sd = 1.05) as more accurate than in divergent ones (mean = 3.40, sd = 1.08; \(\hat{b}_{\text{Accuracy}}\) = 1.88 {[}1.658, 2.102{]}, p = \textless{} .001). As for competence, participants rated informants in convergent scenarios (mean = 5.24, sd = 0.99) as more competent than in divergent ones (mean = 3.61, sd = 1.11; \(\hat{b}_{\text{Competence}}\) = 1.62 {[}1.411, 1.839{]}, p = \textless{} .001).

The second set of hypotheses targeted the interaction of informational dependency and convergence (Fig. \ref{fig:exp3-plot}). In the independence condition, the effect of convergence on accuracy was more positive (\(\hat{b}_{\text{interaction, Accuracy}}\) = 0.99 {[}0.634, 1.348{]}, p = \textless{} .001) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.89 {[}0.636, 1.142{]}, p = \textless{} .001). Likewise the effect of convergence on competence is more positive (\(\hat{b}_{\text{interaction, Competence}}\) = 0.80 {[}0.474, 1.13{]}, p = \textless{} .001) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.82 {[}0.591, 1.056{]}, p = \textless{} .001).

@@

In an exploratory, non-preregistered analysis, we tested whether the effect of convergence is larger on accuracy than on competence. To this end, we regressed the outcome score on convergence and its interaction with a binary variable indicating which outcome was asked for (accuracy or competence), while controlling for informational dependency. We find a negative interaction effect, indicating that pooled across independent and conflict of interest conditions, the effect of convergence had as smaller effect on competence than on accuracy (\(\hat{b}\) = -0.16 {[}-0.294, -0.028{]}, 0.018). Pooled across all conditions, participants reported higher perceived competence than accuracy (\(\hat{b}\) = 0.11 {[}0.025, 0.191{]}, 0.011).

@@

As in the numerical setting, we found a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 16.84 {[}15.009, 18.668{]}, p = \textless{} .001; on a scale from 0 to 100) and competence (\(\hat{b}_{\text{Competence}}\) = 0.68 {[}0.578, 0.788{]}, p = \textless{} .001; on a scale from 1 to 7).

@@

We find evidence for all four hypotheses (see Fig. \ref{fig:exp5-plot}). To test H1a and H1b, we use the same analyses as in Experiment 4, restricted on the independence condition, and replicate the results. We find a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 12.34 {[}10.362, 14.311{]}, p = \textless{} .001) and competence (\(\hat{b}_{\text{Competence}}\) = 0.56 {[}0.459, 0.665{]}, p = \textless{} .001).

The second set of hypotheses targeted the interaction of informational dependency and convergence (Fig. \ref{fig:exp5-plot}). In the independence condition, the effect of convergence on accuracy was more positive (\(\hat{b}_{\text{interaction, Accuracy}}\) = 3.01 {[}0.027, 5.988{]}, p = 0.048) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) 9.33 {[}7.232, 11.426{]}, p = \textless{} .001). Likewise, the effect of convergence on competence was more positive (\(\hat{b}_{\text{interaction, Competence}}\) = 0.16 {[}0.014, 0.316{]}, p = 0.032) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.40 {[}0.291, 0.503{]}, p = \textless{} .001).

@@

We ran a power simulation to inform our choice of sample size. All assumptions and details on the procedure can be found on the OSF. We used previous experiments and estimates of our models to inform our choice of parameter values. We ran two different power analyses, one for each outcome variable. We set the power threshold for our experiment to 90\%. The power simulation for accuracy suggested that for 140 participants we would cross the power threshold of 90\% for the interaction effect (power = 0.928). The simulation for competence suggested that with 300 participants, we would detect an interaction with a power of 87\%. Due to budget constraints, we considered aiming for a sample of 300 participants as good enough, although slightly below our threshold. Due to two failed attention checks, our final sample consisted of 298 subjects, recruited, as in all experiments, in the UK and via Prolific (149 female, 149, 1 non-identified; \(age_\text{mean}\): 42.09, \(age_\text{sd}\): 13.06, \(age_\text{median}\): 40).

@@

We replicate the results of experiment 4, but do not find evidence for an interaction between convergence and the number of choice options. To match the setting of experiment one, we reduced the sample to half of the participants, namely those who were assigned to the three options condition. On this reduced sample, we ran the exact same analyses as in experiment 4 and replicated the results. We find a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 15.68 {[}14.112, 17.246{]}, p = \textless{} .001) and competence (\(\hat{b}_{\text{Competence}}\) = 0.65 {[}0.564, 0.736{]}, p = \textless{} .001). This finding also holds across the entire sample, pooling three and ten choice option observations (\(\hat{b}_{\text{Accuracy}}\) = 16.30 {[}15.124, 17.485{]}, \textless{} .001; \(\hat{b}_{\text{Competence}}\) = 0.68 {[}0.611, 0.739{]}, \textless{} .001).

We do not find evidence of an interaction, i.e.~evidence that the number of choice options changes the effect of convergence (\(\hat{b}_{\text{interaction, Accuracy}}\) = 1.25 {[}-1.11, 3.613{]}, 0.298; \(\hat{b}_{\text{interaction, Competence}}\) = 0.05 {[}-0.078, 0.178{]}, 0.442).

@@

To test the other two research questions, we ran the same mixed-models as for the hypotheses, but this time including an interaction between number of informants and convergence. We use deviation-coded versions of our convergent variable (divergent = -0.5, convergent = 0.5), allowing us to have a coefficient measuring the main effect of the number of informants in our model. Regarding RQ2, pooling across convergent and divergent conditions, we find a main effect of number, such that participants had more confidence in their estimate when they relied on ten informants compared to three informants (\(\hat{b}_{\text{Accuracy}}\) = 0.18 {[}0.076, 0.274{]}, p = \textless{} .001). Regarding RQ3, we find an interaction between number of informants and convergence on competence: the positive effect of convergence was stronger in scenarios involving ten informants compared to three informants (\(\hat{b}_{\text{Competence}}\) = 0.22 {[}0.028, 0.402{]}, p = 0.024). We do not find a statistically significant interaction on accuracy (\(\hat{b}_{\text{Accuracy}}\) = 0.19 {[}-0.007, 0.387{]}, p = 0.059).

>>>>

\caption{Results of the analytic argument. The figure shows the average estimated competence for a focal individual, depending on the proportion of individuals agreeing with them, in a categorical choice scenario. Here, we assume that 20 individuals independently pick one out of 5 choice options. Two situations are represented: one in which there is low variance in the distribution of competence in the population (the bell-shaped curve; solid line), and one in which there is high variance in this distribution (the u-shaped curve; dotted line).}\label{fig:figure-Benoit-2}
\end{figure}

@@

To account for dependencies of observations due to our within-participant design, we ran mixed models, with a random intercept and a random slopes of convergence for participants. In the models for our hypotheses, we control for the number of estimates provided to the participants (three or ten). Visualizations and descriptive statistics can be found in ESM, section \ref{exp1}. We find a positive effect of convergence on accuracy: Participants were more confident about their estimate in convergent scenarios (mean = 4.56, sd = 1.45) than in divergent ones (mean = 3.19, sd = 1.39; \(\hat{b}_{\text{Accuracy}}\) = 1.37 {[}1.225, 1.51{]}, p \textless{} .001). We also find a positive effect of convergence on competence: participants rated players as more competent in convergent scenarios (mean = 4.75, sd = 1.24) than in divergent ones (mean = 3.52, sd = 1.27; \(\hat{b}_{\text{Competence}}\) = 1.23 {[}1.065, 1.4{]}, p \textless{} .001).

In an exploratory, non-preregistered analysis, we tested whether the effect of convergence is larger on accuracy than on competence. To this end, we regressed the outcome score on convergence and its interaction with a binary variable indicating which outcome was asked for (accuracy or competence), while controlling for the number of informants. We do not find a statistically significant interaction that would indicate a difference of the effect of convergence (\(\hat{b}\) = 0.14 {[}-0.001, 0.271{]}, = 0.052). Pooled across divergent and convergent conditions, we find that participants reported lower perceived accuracy than competence (\(\hat{b}\) = -0.26 {[}-0.359, -0.156{]}, \textless{} .001)

@@

To account for dependencies of observations due to our within-participant design, we ran mixed models, with a random intercept and a random slope for convergence for participants. We find evidence for all four hypotheses. As for the first set of hypotheses, to match the setting of experiment one, we reduced the sample of Experiment 3 to half of the participants, namely those who were assigned to the independence condition. On this reduced sample, we ran the exact same analyses as in Experiment 1 and replicated the results. As for accuracy, participants rated informants in convergent scenarios (mean = 5.28, sd = 1.05) as more accurate than in divergent ones (mean = 3.40, sd = 1.08; \(\hat{b}_{\text{Accuracy}}\) = 1.88 {[}1.658, 2.102{]}, p \textless{} .001). As for competence, participants rated informants in convergent scenarios (mean = 5.24, sd = 0.99) as more competent than in divergent ones (mean = 3.61, sd = 1.11; \(\hat{b}_{\text{Competence}}\) = 1.62 {[}1.411, 1.839{]}, p \textless{} .001).

The second set of hypotheses targeted the interaction of informational dependency and convergence (Fig. \ref{fig:exp3-plot}). In the independence condition, the effect of convergence on accuracy was more positive (\(\hat{b}_{\text{interaction, Accuracy}}\) = 0.99 {[}0.634, 1.348{]}, p \textless{} .001) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.89 {[}0.636, 1.142{]}, p \textless{} .001). Likewise the effect of convergence on competence is more positive (\(\hat{b}_{\text{interaction, Competence}}\) = 0.80 {[}0.474, 1.13{]}, p \textless{} .001) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.82 {[}0.591, 1.056{]}, p \textless{} .001).

@@

In an exploratory, non-preregistered analysis, we tested whether the effect of convergence is larger on accuracy than on competence. To this end, we regressed the outcome score on convergence and its interaction with a binary variable indicating which outcome was asked for (accuracy or competence), while controlling for informational dependency. We find a negative interaction effect, indicating that pooled across independent and conflict of interest conditions, the effect of convergence had as smaller effect on competence than on accuracy (\(\hat{b}\) = -0.16 {[}-0.294, -0.028{]}, = 0.018). Pooled across all conditions, participants reported higher perceived competence than accuracy (\(\hat{b}\) = 0.11 {[}0.025, 0.191{]}, = 0.011).

@@

As in the numerical setting, we found a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 16.84 {[}15.009, 18.668{]}, p \textless{} .001; on a scale from 0 to 100) and competence (\(\hat{b}_{\text{Competence}}\) = 0.68 {[}0.578, 0.788{]}, p \textless{} .001; on a scale from 1 to 7).

@@

We find evidence for all four hypotheses (see Fig. \ref{fig:exp5-plot}). To test H1a and H1b, we use the same analyses as in Experiment 4, restricted on the independence condition, and replicate the results. We find a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 12.34 {[}10.362, 14.311{]}, p \textless{} .001) and competence (\(\hat{b}_{\text{Competence}}\) = 0.56 {[}0.459, 0.665{]}, p \textless{} .001).

The second set of hypotheses targeted the interaction of informational dependency and convergence (Fig. \ref{fig:exp5-plot}). In the independence condition, the effect of convergence on accuracy was more positive (\(\hat{b}_{\text{interaction, Accuracy}}\) = 3.01 {[}0.027, 5.988{]}, p = 0.048) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) 9.33 {[}7.232, 11.426{]}, p \textless{} .001). Likewise, the effect of convergence on competence was more positive (\(\hat{b}_{\text{interaction, Competence}}\) = 0.16 {[}0.014, 0.316{]}, p = 0.032) than in the conflict of interest condition (\(\hat{b}_{\text{baseline}}\) = 0.40 {[}0.291, 0.503{]}, p \textless{} .001).

@@

We ran a power simulation to inform our choice of sample size. All assumptions and details on the procedure can be found on the OSF. We used previous experiments and estimates of our models to inform our choice of parameter values. We ran two different power analyses, one for each outcome variable. We set the power threshold for our experiment to 90\%. The power simulation for accuracy suggested that for 140 participants we would cross the power threshold of 90\% for the interaction effect (power = 0.928). The simulation for competence suggested that with 300 participants, we would detect an interaction with a power of 87\%. Due to budget constraints, we considered aiming for a sample of 300 participants as good enough, although slightly below our threshold. Due to two failed attention checks, our final sample consisted of 298 subjects, recruited, as in all experiments, in the UK and via Prolific (149 female, 149; \(age_\text{mean}\): 42.09, \(age_\text{sd}\): 13.06, \(age_\text{median}\): 40).

@@

We replicate the results of experiment 4, but do not find evidence for an interaction between convergence and the number of choice options. To match the setting of experiment one, we reduced the sample to half of the participants, namely those who were assigned to the three options condition. On this reduced sample, we ran the exact same analyses as in experiment 4 and replicated the results. We find a positive effect of convergence on both accuracy (\(\hat{b}_{\text{Accuracy}}\) = 15.68 {[}14.112, 17.246{]}, p \textless{} .001) and competence (\(\hat{b}_{\text{Competence}}\) = 0.65 {[}0.564, 0.736{]}, p \textless{} .001). This finding also holds across the entire sample, pooling three and ten choice option observations (\(\hat{b}_{\text{Accuracy}}\) = 16.30 {[}15.124, 17.485{]}, \textless{} .001; \(\hat{b}_{\text{Competence}}\) = 0.68 {[}0.611, 0.739{]}, \textless{} .001).

We do not find evidence of an interaction, i.e.~evidence that the number of choice options changes the effect of convergence (\(\hat{b}_{\text{interaction, Accuracy}}\) = 1.25 {[}-1.11, 3.613{]}, = 0.298; \(\hat{b}_{\text{interaction, Competence}}\) = 0.05 {[}-0.078, 0.178{]}, = 0.442).

@@

To test the other two research questions, we ran the same mixed-models as for the hypotheses, but this time including an interaction between number of informants and convergence. We use deviation-coded versions of our convergent variable (divergent = -0.5, convergent = 0.5), allowing us to have a coefficient measuring the main effect of the number of informants in our model. Regarding RQ2, pooling across convergent and divergent conditions, we find a main effect of number, such that participants had more confidence in their estimate when they relied on ten informants compared to three informants (\(\hat{b}_{\text{Accuracy}}\) = 0.18 {[}0.076, 0.274{]}, p = \textless{} .001). Regarding RQ3, we find an interaction between number of informants and convergence on competence: the positive effect of convergence was stronger in scenarios involving ten informants compared to three informants (\(\hat{b}_{\text{Competence}}\) = 0.22 {[}0.028, 0.402{]}, p = = 0.024). We do not find a statistically significant interaction on accuracy (\(\hat{b}_{\text{Accuracy}}\) = 0.19 {[}-0.007, 0.387{]}, p = = 0.059).
